---
title: "RNA-seq Data"
author: Thomas Walsh
date: March 30, 2026
format: 
  html:
    theme: 
      light: flatly
      dark: darkly
execute:
  echo: false
editor: visual
---

```{r}
#| include: false
#| message: false
library(readr) # install.packages("readr")
library(dplyr)
library(tidyr)

library(ggplot2)
library(cowplot)
library(hexbin)  # install.packages("hexbin")
theme_set(theme_cowplot())
library(RColorBrewer)
library(here) # install.packages("here")

if(FALSE){
    download.file(url = "https://github.com/Bioconductor/bioconductor-teaching/raw/master/data/GSE96870/rnaseq.csv",
              destfile = file.path(here::here(), "data", "rnaseq.csv"))
}

### Read in data here ###
dat <- read_csv(file.path(here::here(), "data", "rnaseq.csv")) |>
    mutate(l2exp = log2(expression + 1))
```

## Agenda

- Today's Exercise
  - RNA-Seq data visualization - this lab borrows heavily from a [Carpentries incubator module on visualization](https://carpentries-incubator.github.io/bioc-intro/40-visualization.html).

## RNA-Seq data

To start out with, fork the starter code and clone the repository to your local machine. Once we have done that, let's explore the data provided:

```{r}
dat
```

Take a look at the distribution of the expression levels:

```{r}
#| message: false
# () around the 'g1' variable enables you to create it and plot it in one call.
(g1 <- dat |>
    ggplot(expression) +
    geom_histogram(bins = 50))
# You'll probably notice an automatic message that appears when drawing the histogram: `stat_bin()` using `bins = 30`. Pick better value with `binwidth`.
# Change the arguments bins or binwidth of geom_histogram() to change the number or width of the bins.
```

The data should be transformed. Transforming on the $\log_{10}$ scale is easy in `ggplot2`,

```{r}
#| warning: false
# this just scales fig 'g1' to log 10
g1 + scale_x_log10()
```

but we often use a $log_2$ scale when looking at expression data:

```{r}
#| warning: false
ticks_at <- c(0,5,10,15)
ggplot(dat, aes(l2exp)) +
    geom_histogram(bins = 50) +
    scale_x_continuous(labels = 2^ticks_at,
                        breaks = ticks_at) +
    labs(x = "Expression")
```

### Building plots iteratively

We will now draw a scatter plot with two continuous variables and the `geom_point()` function. This graph will represent the $\log_2$ fold changes of expression comparing time 8 versus time 0, and time 4 versus time 0. To this end, we first need to compute the means of the log-transformed expression values by gene and time, then the log fold changes by subtracting the mean log expressions between time 8 and time 0 and between time 4 and time 0.

```{r}
#| message: false
# fold change data
dat_fc <- dat |> 
    select(gene, 
           time,
           gene_biotype,
           expression) |>
    
     # calculate summaries of log2 expression grouped by gene, time, gene_biotype
    group_by(gene, time, gene_biotype) |>
    summarize(mean_exp = mean(log2(expression + 1))) |>
    ungroup() |>
    
    # pivot wider so that there is a separate column for each value in time
    pivot_wider(names_from = time, 
                values_from = mean_exp,
                names_prefix = 't') |>

    mutate(time_8_vs_0 = `t8` - `t0`, # compare time 8 vs 0
           time_4_vs_0 = `t4` - `t0`) # compare time 4 vs 0

# visualize time_4_vs_0 by time_8_vs_0
ggplot(dat_fc, aes(time_4_vs_0, time_8_vs_0)) +
    geom_point(alpha = .33) +
    geom_smooth(method = 'lm', 
                color = 'orange',
                se = FALSE,
                linewidth = .6)
```

We could also add color, but it begins to get a little messy with the large number of categories we have.

```{r}
ggplot(dat_fc, aes(time_4_vs_0, time_8_vs_0, color = gene_biotype)) +
    geom_point(alpha = .33) +
        geom_smooth(method = 'lm', 
                    color = 'black',
                    se = FALSE,
                    linewidth = .6)

```

Over-plotting can be an issue with large datasets like this. One solution is using `hexbin:geom_hex()`.

```{r}
ggplot(dat_fc, aes(time_4_vs_0, time_8_vs_0)) +
    geom_hex() +
    geom_smooth(method = 'lm', 
                    color = 'black',
                    se = FALSE,
                    linewidth = .6)

```

### Box plots

We can use boxplots to visualize the distribution of gene expressions within each sample:

```{r}
(g2 <- ggplot(dat, aes(sample, l2exp)) +
        geom_boxplot() +
        theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
        labs(y = "Log2 Expression", x = "Sample"))

```

Adding points to box plots can give us a better understanding of the underlying distributions.

```{r}
g2 + geom_jitter(alpha = 0.02, width = .15)

```

We could also use `geom_violin` to better see the distribution of points. Try coloring the points by `time` as well.

```{r}
ggplot(dat, aes(sample, l2exp)) +
    geom_violin() +
    geom_jitter(alpha = 0.02, width = .15, aes(color = as.factor(time))) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    labs(y = "Log2 Expression", x = "Sample", color = "Time")

```

### Line plots

Let's calculate the mean expression per duration of the infection for the 10 genes having the highest log fold changes comparing time 8 versus time 0. We can do this either by coloring each gene or using `facet_wrap`.

```{r}
#| message: false

# First, we need to select the genes
genes_selected <- arrange(dat_fc, desc(time_8_vs_0))$gene[1:10]

# Then create a subset of dat containing the 10 selected genes
mean_exp_by_time <- dat |>
    filter(gene %in% genes_selected) |>
    
    # Then group the data by gene and time
    group_by(gene, time) |>
    # calculate the mean gene expression within each group
    summarize(mean_exp = mean(l2exp)) |>
    ungroup()

ggplot(mean_exp_by_time, 
        aes(time, mean_exp, 
            group = gene, color = gene)) +
    geom_line() +
    labs(y = "Mean Log 2 Expression", x = "Time", color = "Gene")

# similar, but w/out color and w/facet wrap for multi.plots
ggplot(mean_exp_by_time, 
        aes(time, mean_exp)) +
    geom_line() +
    facet_wrap(~ gene) +
    labs(y = "Mean Log 2 Expression", x = "Time")

```

Now we would like to split the line in each plot by the sex of the mice.

```{r}
#| message: false
# Then create a subset of dat containing the 10 selected genes
mean_exp_by_time_sex <- dat |>
    filter(gene %in% genes_selected) |>
    
    # Then group the data by gene, time, and sex
    group_by(gene, time, sex) |>
    # calculate the mean gene expression within each group
    summarize(mean_exp = mean(l2exp)) |>

ggplot(mean_exp_by_time_sex,
        aes(time, mean_exp, group = sex, color = sex)) +
    geom_line() +
    facet_wrap(~ gene) +
    labs(y = "Mean Log 2 Expression", x = "Time", color = "Sex") +
    scale_color_brewer(palette = "Dark2")
```

The `facet_wrap` geometry extracts plots into an arbitrary number of dimensions to allow them to cleanly fit on one page. On the other hand, the `facet_grid` geometry allows you to specify variables to plot by row or by column (see `?facet_grid`).

Let's modify the previous plot to compare how the mean gene expression of males and females has changed through time:

```{r}
# One column, facet by rows
ggplot(mean_exp_by_time_sex, 
        aes(time, mean_exp,
            group = gene, color = gene,)) +
    geom_line() +
    facet_grid(sex ~ .) +
    labs(y = "Mean Log 2 Expression", x = "Time", color = "Gene")
```